The dynamics of starvation and recovery

### 2016-08-09 Manuscript/8.tex
Manuscript and body size invasion stuff edits JY

The life history of most species is typically comprised of (a) somatic growth
and maintenance, and (b) reproduction.  The balance between these two
activities is often conditioned on resource
availability~\cite{Morris:1987eo}.  For example, reindeer invest less in
calves born after harsh winters (when the mother's energetic state is
depleted) than in calves born after moderate winters~\cite{Tveraa:2003fq}.
Many bird species invest differently in broods during periods of resource
scarcity compared to normal periods~\cite{Daan:1988va,Jacot:2009dv},
sometimes delaying or even foregoing reproduction for a breeding
season~\cite{Martin:1987dl,Stearns:1989ip,Barboza:2002in}.  Even freshwater
and marine zooplankton have been observed to avoid reproduction under
nutritional stress~\cite{Threlkeld:1976ih}, and those that do reproduce have
lower survival rates~\cite{Kirk:1997cc}.  Organisms may also separate
maintenance and growth from reproduction over space and time: many salmonids,
birds, and some mammals return to migratory breeding grounds to reproduce
after one or multiple seasons in resource-rich environments where they
accumulate nutritional
reserves~\cite{Weber:1998jg,Mduma:1999cp,Moore:2014hi}.

Physiological mechanisms also play an important role in regulating
reproductive expenditures during periods of resource limitation.  The data
collected thus far has shown that diverse mammals (47 species in 10 families)
exhibit delayed implantation, whereby females postpone fetal development
(blastocyst implantation) until times where nutritional reserves can be
accumulated~\cite{Mead:1989dt,Sandell:1990kw}.  Many other many species
(including humans) suffer irregular menstrual cycling and higher spontaneous
abortion rates during periods of nutritional
stress~\cite{Bulik:1999eo,Trites:2003cc}.  In the extreme case of unicellular
organisms, nutrition is unavoidably linked to reproduction because the
nutritional state of the cell regulates all aspects of the cell cycle
\cite{Glazier:2009hq}.  The existence of so many independently evolved
mechanisms across such a diverse suite of organisms highlights the importance
and universality of the fundamental tradeoff between somatic and reproductive
investment.  However the dynamic implications of these constraints are
unknown.

@@

\noindent {\bf Nutritional-state-structured model (NSM)}\\
We begin by defining a minimal Nutritional-State-structured population Model
(NSM), where the consumer population is divided into two energetic states:
(a) an energetically replete (full) state $F$, where the consumer reproduces
at a constant rate $\lambda$ and has no mortality risk, and (b) an
energetically deficient (hungry) state $H$, where the consumer does not
reproduce but dies at rate $\mu$.  The underlying resource $R$ evolves by
logistic growth with an intrinsic growth rate $\alpha$ and a carrying
capacity equal to one.  Consumers transition from the full state $F$ to the
hungry state $H$ at a rate $\sigma$---the starvation rate---and also in
proportion to the absence of resources $(1-R)$.  Conversely, consumers
recover from state $H$ to state $F$ at rate $\rho$ and in proportion to $R$.
Resources are also eaten by the consumers---at rate $\rho$ by hungry
consumers and at rate $\beta<\rho$ by full consumers.  This inequality
accounts for hungry consumers requiring more resources to rebuild body
weight.

@@

Nearly all of the rates described in the NSM are to some extent governed by
consumer metabolism, which can be used to describe a variety of organismal
features \cite{Brown:2004wq}.  The scaling relation between an organism's
metabolic rate $B$ and its body size at reproductive maturity $M$ is well
documented~\cite{West:2002it} and scales as $B = B_0 M^\eta$, where $\eta$ is
the scaling exponent, generally assumed to vary around $2/3$ or $3/4$ for
metazoans (e.g. \cite{Brown:2004wq}), and has taxonomic shifts for unicellular species
between $\eta\approx 1$ in eukaryotes and $\eta\approx 1.76$ in bacteria
\cite{DeLong:2010dy,Kempes:2012hy}. Several efforts have shown how a
partitioning of this metabolic rate between growth and maintenance purposes
can be used to derive a general equation for the growth trajectories and
growth rates of organisms ranging from bacteria to metazoans
\cite{West:2001bv,moses2008rmo,gillooly2002esa,hou,Kempes:2012hy}. More specifically, the cross-species trends in growth rate can be
approximated by
\begin{eqnarray}
\lambda = \lambda_0 M^{\eta-1}.
\label{lambda}
\end{eqnarray}
This relation is derived from the simple balance condition \cite{West:2001bv,moses2008rmo,gillooly2002esa,hou,Kempes:2012hy}
\begin{eqnarray}
\label{balance}
B_{0}m^{\eta}=E_{m}\frac{dm}{dt}+B_{m}m\,,
\end{eqnarray}
where $E_{m}$ is the energy needed to synthesize a unit of mass, $B_{m}$ is
the metabolic rate to support an existing unit of mass, and $m$ is the mass
at any point in development.  This balance has the general solution \cite{moses2008rmo, Kempes:2012hy}
\begin{eqnarray}
m\left(t\right)\!=\left[1\!-\!\left(1\!-\!\frac{b}{a}m_{0}^{1\!-\!\eta}\right)e^{-b\left(1\!-\!\eta\right)t}\right]^{1/\left(1-\eta\right)}\!\left(\!\frac{a}{b}\!\right)^{1/(1-\eta)}
\end{eqnarray}
where, $a=B_{0}/E_{m}$ and $b=B_{m}/E_{m}$ , andwhich we use to define the timescale of recovery from starvation. The rate of recovery $\rho = 1/t_\rho$ requires that an organism accrues
sufficient tissue to transition from the hungry to the full state.  Since only certain tissues can be digested for energy (for example the brain
cannot be degraded to fuel metabolism), we define the rate for starvation,
death, and recovery by the timescales required to reach, or return from, specific fractions of the
replete-state mass. We define $m_{\rm starve}=\epsilon M$, where
$\epsilon<1$ is the fraction of replete-state mass where reproduction ceases. This fraction will be modified if tissue composition systematically scales
with adult mass.  For example, making use of the observation that body fat in
mammals scales with overall body size according to $M_{f}=f_{0}M^{\gamma}$
and assuming that once this mass is fully digested the organism starves, this
would imply that $\epsilon=1-f_{0}M^{\gamma}/M$. It follows that the recovery timescale, $t_{\rho}=t_{2}-t_{1}$, is the time it takes
to go from $m\left(t_{1}\right)=\epsilon M$ to $m\left(t_{2}\right)=M$, or
\begin{eqnarray}
t_{\rho}=\frac{\ln\left[1\!-\!\frac{a}{b}\left(M
  \right)^{1\!-\!\eta }\right]
-\ln \left[1\!-\!\frac{a}{b}\left(\epsilon M \right)^{1\!-\!\eta }\right]}{(\eta -1) b}.
\end{eqnarray}
It should be noted that more complicated ontogenetic models explicitly handle storage \cite{hou}, whereas this is implicitly covered by the body fat scaling in our framework.

To determine the starvation rate, $\sigma$, we are interested in the time
required for an organism to go from a mature adult that reproduces at rate
$\lambda$ (henceforth we term this state as the ``replete'' state), to a
reduced-mass hungry state where reproduction is impossible.  For starving individuals we assume that an organism must meet its maintenance requirements using the digestion of existing mass as the sole energy source.
This assumption implies the following simple metabolic balance 
\begin{eqnarray}
\frac{dm}{dt}E_{m}^{\prime}=-B_{m}m
\end{eqnarray}
where $E_{m}^{\prime}$ is the amount of energy stored in a unit of existing
body mass which differs from $E_{m}$, the energy required to
synthesis a unit of biomass \cite{hou}. Given the replete mass, $M$, of an organism, the
above energy balance prescribes the mass trajectory of a non-consuming
organism:
\begin{eqnarray}
\label{mt}
m\left(t\right)=Me^{-b^{\prime}t}
\end{eqnarray}
where $b^{\prime}=E_{m}^{\prime}/B_{m}$. The time time scale for starvation is given by the time it takes this decay to reach $\epsilon M$, which is
\begin{equation}
\label{eq:sigma}
t_{\sigma}=-b^{\prime}\ln\left(\epsilon\right).
\end{equation}
The starvation rate is then $\sigma=1/t_{\sigma}$, which scales with
replete-state mass as $1/\ln\left(1-f_{0}M^{\gamma}/M\right)$.  An important
feature is that $\sigma$ does not have a simple scaling dependence on
$\lambda$ (Eq.~\ref{lambda}), which is important for the dynamics that we
later discuss.

The time to death should follow a similar relation, but defined by a lower
fraction of replete-state mass, $m_{\rm death}=\epsilon^{\prime} M$.
Suppose, for example, that an organism dies once it has digested all fat and
muscle tissues, and that muscle tissue scales with body mass according to
$M_{m}=u_{0}M^{\zeta}$.  This gives
$\epsilon^{\prime}=1-\left(f_{0}M^{\gamma}+u_{0}M^{\zeta}\right)/M$. Muscle
mass has been shown to be roughly proportional to body mass~\cite{muscle} in
mammals and thus $\epsilon^{\prime}$ is merely $\epsilon$ minus a constant.
Thus
\begin{eqnarray}
t_{\mu}=-b^{\prime}\ln\left(\epsilon^{\prime}\right)
\end{eqnarray}
and $\mu=1/t_{\mu}$.

@@

\noindent {\bf  Stabilizing effects of allometric constraints} \\
As the allometric derivations of the NSM rate laws reveal, $\sigma$ and
$\rho$ are not independent parameters, and the bifurcation space shown in
Fig.~\ref{fig:fp} is navigated via covarying parameters.  Given the
parameters of terrestrial endotherms, we find that $\sigma$ and $\rho$ are
constrained to lie within a small window of potential values
(Fig.~\ref{fig:hopf}) for the known range of body sizes $M$.  We thus find
that the dynamics for all mammalian body sizes is confined to the
steady-state regime of the NSM and that  limit-cycle behavior is precluded.
%, which for mammals ranges from ca. 1 gram (the Etruscan shrew) to ca. $10^7$ grams (represented by the Indricotheriinae, a subfamily of mammals living from the mid-Eocene to early Miocene).
Moreover, for larger $M$, the distance to the Hopf bifurcation increases,
while uncertainty in allometric parameters (20\% variation around the mean;
Fig.~\ref{fig:hopf}) results in little qualitative difference in the distance
to the the Hopf bifurcation. These results suggest that small mammals are
more prone to population oscillations---including both stable limit cycles as
well as transient cycles---than mammals with larger body size.  Thus our NSM
model predicts that population cycles should be less common for larger
species and more common for smaller species, particularly in environments
where resources are limiting.%, and
%2) the decreasing distance to the Hopf bifurcation with lower body sizes is suggestive of a dynamic barrier to the mass of endothermic organisms.

Previous studies have used allometric constraints to explain the periodicity
of cyclic populations
\cite{CalderIII:1983jd,Peterson:1984hj,Krukonis:1991fk}, suggesting a period
$\propto M^{0.25}$, however this relation seems to hold only for some species
\cite{Hendriks:2012fc} and competing explanations [related to ....\sid{???}]
exist~\cite{Kendall:1999iy,Hogstedt:2005cr}.
%It is more difficult to ascertain whether a population is truly cycling according to an internal or external dynamic~\cite{Kendall:1999iy} or if fluctuations are due to stochasticity, and this is particularly true for longer-lived organisms.
Statistically significant support for the existence of population cycles
among mammals is predominantly based on time series for small
mammals~\cite{Kendall:1998hl}, where we our model would predict much longer
and more pronounced transient dynamics, given how close these points are to
the Hopf bifurcation.  On the other hand, the longer gestational times and
the increased difficulty in measurements, precludes obtaining similar-quality
data for larger organisms.
\vspace{6mm}
%, we acknowledge that absence of evidence for cyclic dynamics among longer-lived species is not necessarily evidence of absence.
 %, emphasizing the role of starvation on dynamics.

%Extinction to the left and the right
\noindent {\bf Extinction risk}\\
Within our model, higher rates of starvation result in a larger flux of the
population to the hungry state.  In this state reproduction is absent, thus
increasing the likelihood of extinction.  However, from the perspective of
population survival, it is the rate of starvation relative to the rate of
recovery that determines the long-term dynamics of the system
(Fig.~\ref{fig:fp}).  We now examine the competing effects of cyclic dynamics
vs.\ changes in steady state density on extinction risk as a function of the
ratio $\sigma/\rho$.  To this end, we computed the probability of extinction,
where extinction is defined as the population trajectory going below
$0.2~\times$ the allometrically constrained steady state for all times
between $10^2$ and $\leq 10^6$.  This procedure is repeated for 1000
replicates of the continuous-time system shown in Eq.~\ref{eq:system} for an
organism of $M=100$ grams.  
In each replicate the initial condition is
distributed around the steady state (Eq.~\ref{eq:ss}).  Specifically the
initial densities are chosen to be $A(F^*,H^*,R^*)$, with $A$ a random
variable that is uniformly distribution in $[0,2]$.  By allowing the rate of
starvation to vary, we assessed extinction risk across a range of values of
the ratio $\sigma/\rho$ varying between $10^{-2}$ to $2.5$, thus examining a
horizontal cross-section of Fig.~\ref{fig:fp}.  
As expected, higher rates of
extinction correlated with both low and high values of $\sigma/\rho$.  For
low values of $\sigma/\rho$, the increased extinction risk results from
transient cycles with larger amplitudes as the system nears the Hopf
bifurcation (Fig.~\ref{fig:ext}).  For large values of $\sigma/\rho$, higher
extinction risk arises because of to the decrease in the steady state
consumer population density.  This interplay creates an `extinction refuge'
as shown in Fig.~\ref{fig:ext}, such that for a relatively constrained range
of $\sigma/\rho$, extinction probabilities are minimized.

We find that the allometrically constrained values of $\sigma/\rho$ (with
$\pm 20\%$ variability around energetic parameter means) fall within the
extinction refuge. These values are close enough to the Hopf bifurcation to
avoid low steady state densities, and far enough away to avoid
large-amplitude transient cycles. The fact that allometric values of $\sigma$
and $\rho$ fall within this relatively small window supports the possibility
that a selective mechanism has constrained the physiological conditions that
drive observed starvation and recovery rates within populations.  Such a
mechanism would select for organism physiology that generates appropriate
$\sigma$ and $\rho$ values that avoid extinction.  This selection could occur
via the tuning of body fat percentages, metabolic rates, and biomass
maintenance efficiencies.
% OLD WORDING Such a mechanism would involve a feedback between the dynamics of the population and the fitness of individuals within the population, though to what extent the dynamics of the population influence rates of starvation and recovery would also involve potential tradeoffs in reproduction and somatic maintenance.
To summarize, our finding that the allometrically-determined parameters fall within this low extinction probability region suggests that the NSM dynamics may both drive---and constrain---natural animal populations.\\

@@

%Minimal mass bound
A lower bound on mammalian body size is given by $\epsilon=1$, where mammals
have no metabolic reserves and immediately starve; this occurs at a size of
\sid{$M=value$}.  This calculation \sid{what calculation?} gives an extreme
limit on size but does not account for the subtleties of starvation dynamics
that may limit body size. The NSM correctly predicts that species with
smaller masses have larger steady-state population densities.  However we
observe that there is a sharp change in the mass dependence of both the
steady-state densities and $\sigma/\rho$ at $M \approx 0.3$ grams
(Fig.~\ref{fig:mass}a,b).  The dependence of the rates of starvation and
recovery explain this phenomenon.  As the mass decreases, the rate of
starvation increases, while the rate of recovery declines super-exponentially
\sid{how do we know this?}.  This decline in $\rho$ occurs when the
percentage of body fat is
$1 - \big(a/b\big)^{-1/\left(\eta-1\right)}M^{-1} \approx
2\%$,
whereupon consumers have no eligible route \sid{what does this mean?} that avoids
starvation.  Compellingly, this dynamic bound determined by the rate of
energetic recovery is close to the minimum observed mammalian body size of
ca. 1.3-2.5 grams (Fig.~\ref{fig:mass}b,c), a range that occurs as the
recovery rate begins its decline.  In addition to known transport
limitations~\cite{West:2002ud}, we suggest that an additional constraint of
lower body size stems from the dynamics of starvation.  This work \sid{which
  work?} mirrors other efforts where coincident limitations seem to limit the
smallest possibilities for life within a particular class or
organisms~\cite{Kempes:2012hy}.

%Maximum mass bound
%Although there are upper bounds to the rate equations (e.g. when percent body fat becomes unity), they are not biologically feasible and we do not discuss them further.
We determine a potential upper bound to body mass by assessing the
susceptibility of an otherwise homogeneous population to invasion by a
mutated subset of the population (denoted by ${}^\prime$) where individuals
have a modified proportion of body fat $M^\prime=M(1+\chi)$ where
$\chi \in [-0.5,0.5]$, thus altering the rates of starvation $\sigma$,
recovery $\rho$, and maintenance $\beta$.  There is no internal fixed point
that correspond to a state where both original residents and invaders coexist
(except for the trivial state $\chi=0$).  To assess the susceptibility to
invasion as a function of the invader mass, we determine which consumer has a
higher steady-state density for a given value of $\chi$.  We find that for
$1\leq M<10^6$g, having additional body fat ($\chi > 0$) results in a higher
steady-state invader population density
($H^{\prime *}+F^{\prime *}>H^*+F^*$).  Thus the invader has an intrinsic
advantage over the resident population.  However, for $M>10^6$, leaner
individuals ($\chi < 0$) have the advantage, and this is due to the changing
covariance between energetic rates as a function of modified energetic
reserves \sid{I don't understand the phrase after the comma}.

>>>>

The balance between (a) somatic growth and maintenance, and (b) reproduction is often conditioned on resource
availability~\cite{Morris:1987eo}.  For example, reindeer invest less in
calves born after harsh winters (when the mother's energetic state is
depleted) than in calves born after moderate winters~\cite{Tveraa:2003fq}.
Many bird species invest differently in broods during periods of resource
scarcity compared to normal periods~\cite{Daan:1988va,Jacot:2009dv},
sometimes delaying or even foregoing reproduction for a breeding
season~\cite{Martin:1987dl,Stearns:1989ip,Barboza:2002in}.  Even freshwater
and marine zooplankton have been observed to avoid reproduction under
nutritional stress~\cite{Threlkeld:1976ih}, and those that do reproduce have
lower survival rates~\cite{Kirk:1997cc}. Organisms may also separate
maintenance and growth from reproduction over space and time: many salmonids,
birds, and some mammals return to migratory breeding grounds to reproduce
after one or multiple seasons in resource-rich environments where they
accumulate nutritional
reserves~\cite{Weber:1998jg,Mduma:1999cp,Moore:2014hi}.

Physiological mechanisms also play an important role in regulating
reproductive expenditures during periods of resource limitation.  The data
collected thus far has shown that diverse mammals (47 species in 10 families)
exhibit delayed implantation, whereby females postpone fetal development
(blastocyst implantation) until nutritional reserves can be
accumulated~\cite{Mead:1989dt,Sandell:1990kw}.  Many other many species
(including humans) suffer irregular menstrual cycling and higher spontaneous
abortion rates during periods of nutritional
stress~\cite{Bulik:1999eo,Trites:2003cc}.  In the extreme case of unicellular
organisms, nutrition is unavoidably linked to reproduction because the
nutritional state of the cell regulates all aspects of the cell cycle
\cite{Glazier:2009hq}.  The existence of so many independently evolved
mechanisms across such a diverse suite of organisms highlights the importance
and universality of the fundamental tradeoff between somatic and reproductive
investment.  However the dynamic implications of these constraints are
unknown.

@@

\noindent {\bf Nutritional state-structured model (NSM)}\\
We begin by defining a minimal Nutritional State-structured population Model
(NSM), where the consumer population is divided into two energetic states:
(a) an energetically replete (full) state $F$, where the consumer reproduces
at a constant rate $\lambda$ and does not die from starvation, and (b) an
energetically deficient (hungry) state $H$, where the consumer does not
reproduce but dies by starvation at rate $\mu$.  The underlying resource $R$ evolves by
logistic growth with an intrinsic growth rate $\alpha$ and a carrying
capacity equal to one.  Consumers transition from the full state $F$ to the
hungry state $H$ at a rate $\sigma$---the starvation rate---and also in
proportion to the absence of resources $(1-R)$.  Conversely, consumers
recover from state $H$ to state $F$ at rate $\rho$ and in proportion to $R$.
Resources are also eaten by the consumers---at rate $\rho$ by hungry
consumers and at rate $\beta<\rho$ by full consumers.  This inequality
accounts for hungry consumers requiring more resources to replace lost body
tissue.

@@

Nearly all of the rates described in the NSM are to some extent governed by
consumer metabolism, which can be used to describe a variety of organismal
features \cite{Brown:2004wq}.  The scaling relation between an organism's
metabolic rate $B$ and its body size at reproductive maturity $m=M$ (where $m$ denotes body sizes during growth and prior to reproductive maturity) is well
documented~\cite{West:2002it} and scales as $B = B_0 M^\eta$, where $\eta$ is
the scaling exponent, generally assumed to vary around $2/3$ or $3/4$ for
metazoans (e.g. \cite{Brown:2004wq}), and has taxonomic shifts for unicellular species
between $\eta\approx 1$ in eukaryotes and $\eta\approx 1.76$ in bacteria
\cite{DeLong:2010dy,Kempes:2012hy}. 
%Justin adding an explanation for beta
An organism's metabolic rate is proportional to the cost of tissue maintenance without growth (i.e. at $m=M$) such that $\beta = \xi B$, where $\xi$ is related to the conversion efficiency of resource to consumer tissue (Supporting Information, section XX).

Several efforts have shown how a
partitioning of $B$ between growth and maintenance purposes
can be used to derive a general equation for the growth trajectories and
growth rates of organisms ranging from bacteria to metazoans
\cite{West:2001bv,moses2008rmo,gillooly2002esa,hou,Kempes:2012hy}. More specifically, the cross-species trends in growth rate can be
approximated by
\begin{eqnarray}
\lambda = \lambda_0 M^{\eta-1}.
\label{lambda}
\end{eqnarray}
This relation is derived from the simple balance condition \cite{West:2001bv,moses2008rmo,gillooly2002esa,hou,Kempes:2012hy}
\begin{eqnarray}
\label{balance}
B_{0}m^{\eta}=E_{m}\dot{m}+B_{m}m\,,
\end{eqnarray}
where $E_{m}$ is the energy needed to synthesize a unit of mass, $B_{m}$ is
the metabolic rate to support an existing unit of mass, and $m$ is the mass
at any point in development.  This balance has the general solution \cite{moses2008rmo, Kempes:2012hy}
\begin{eqnarray}
m\left(t\right)\!=\left[1\!-\!\left(1\!-\!\frac{b}{a}m_{0}^{1\!-\!\eta}\right)e^{-b\left(1\!-\!\eta\right)t}\right]^{1/\left(1-\eta\right)}\!\left(\!\frac{a}{b}\!\right)^{1/(1-\eta)}
\end{eqnarray}
where, $a=B_{0}/E_{m}$, $b=B_{m}/E_{m}$, and $m_0$ is mass at birth, which we use to define the timescale of recovery from starvation. The rate of recovery $\rho = 1/t_\rho$ requires that an organism accrues
sufficient tissue to transition from the hungry to the full state.  Since only certain tissues can be digested for energy (for example the brain
cannot be degraded to fuel metabolism), we define the rate for starvation,
death, and recovery by the timescales required to reach, or return from, specific fractions of the
replete-state mass. 
Here and henceforth, see Supporting Information, sections XX-XX for details on parameterizations.
We define $m_{\rm starve}=\epsilon M$, where
$\epsilon<1$ is the fraction of replete-state mass where reproduction ceases. This fraction will be modified if tissue composition systematically scales
with adult mass.  For example, making use of the observation that body fat in
mammals scales with overall body size according to $M_{\rm fat}=f_{0}M^{\gamma}$
and assuming that once this mass is fully digested the organism starves, this
would imply that $\epsilon=1-f_{0}M^{\gamma}/M$. It follows that the recovery timescale, $t_{\rho}=t_{2}-t_{1}$, is the time it takes
to go from $m\left(t_{1}\right)=\epsilon M$ to $m\left(t_{2}\right)=M$, or
\begin{eqnarray}
t_{\rho}=\frac{\ln\left[1\!-\!\frac{a}{b}\left(M
  \right)^{1\!-\!\eta }\right]
-\ln \left[1\!-\!\frac{a}{b}\left(\epsilon M \right)^{1\!-\!\eta }\right]}{(\eta -1) b}.
\end{eqnarray}
It should be noted that more complicated ontogenetic models explicitly handle storage \cite{hou}, whereas this is implicitly covered by the body fat scaling in our framework.

To determine the starvation rate, $\sigma$, we are interested in the time
required for an organism to go from a mature adult that reproduces at rate
$\lambda$, to a
reduced-mass hungry state where reproduction is impossible.  For starving individuals we assume that an organism must meet its maintenance requirements using the digestion of existing mass as the sole energy source.
This assumption implies the following simple metabolic balance 
\begin{eqnarray}
\dot{m}E_{m}^{\prime}=-B_{m}m
\end{eqnarray}
where $E_{m}^{\prime}$ is the amount of energy stored in a unit of existing
body mass which differs from $E_{m}$, the energy required to
synthesis a unit of biomass \cite{hou}. Given the replete mass, $M$, of an organism, the
above energy balance prescribes the mass trajectory of a non-consuming
organism:
\begin{eqnarray}
\label{mt}
m\left(t\right)=Me^{-b^{\prime}t}
\end{eqnarray}
where $b^{\prime}=E_{m}^{\prime}/B_{m}$. The time scale for starvation is given by the time it takes this decay to reach $\epsilon M$, which is
\begin{equation}
\label{eq:sigma}
t_{\sigma}=-b^{\prime}\ln\left(\epsilon\right).
\end{equation}
The starvation rate is then $\sigma=1/t_{\sigma}$, which scales with
replete-state mass as $1/\ln\left(1-f_{0}M^{\gamma}/M\right)$.  An important
feature is that $\sigma$ does not have a simple scaling dependence on
$\lambda$ (Eq.~\ref{lambda}), which is important for the dynamics that we
later discuss.

The time to death should follow a similar relation, but defined by a lower
fraction of replete-state mass, $m_{\rm death}=\epsilon^{\prime} M$.
Suppose, for example, that an organism dies once it has digested all fat and
muscle tissues, and that muscle tissue scales with body mass according to
$M_{\rm musc}=u_{0}M^{\zeta}$.  This gives
$\epsilon^{\prime}=1-\left(f_{0}M^{\gamma}+u_{0}M^{\zeta}\right)/M$. Muscle
mass has been shown to be roughly proportional to body mass~\cite{Folland:2008ij} in
mammals and thus $\epsilon^{\prime}$ is merely $\epsilon$ minus a constant.
Thus
\begin{eqnarray}
t_{\mu}=-b^{\prime}\ln\left(\epsilon^{\prime}\right),
\end{eqnarray}
and $\mu=1/t_{\mu}$.

@@

\noindent {\bf  Stabilizing effects of allometric constraints} \\
As the allometric derivations of the NSM rate laws reveal, starvation and recovery rates are not independent parameters, and the bifurcation space shown in
Fig.~\ref{fig:fp} is navigated via covarying parameters.  Given the
parameters of terrestrial endotherms, we find that $\sigma$ and $\rho$ are
constrained to lie within a small window of potential values
(Fig.~\ref{fig:hopf}) for the known range of body sizes $M$.  We thus find
that the dynamics for all mammalian body sizes is confined to the
steady-state regime of the NSM and that  limit-cycle behavior is precluded.
%, which for mammals ranges from ca. 1 gram (the Etruscan shrew) to ca. $10^7$ grams (represented by the Indricotheriinae, a subfamily of mammals living from the mid-Eocene to early Miocene).
Moreover, for larger $M$, the distance to the Hopf bifurcation increases,
while uncertainty in allometric parameters (20\% variation around the mean;
Fig.~\ref{fig:hopf}) results in little qualitative difference in the distance
to the the Hopf bifurcation. These results suggest that small mammals are
more prone to population oscillations---including both stable limit cycles as
well as transient cycles---than mammals with larger body size.  Thus our NSM
model predicts that population cycles should be less common for larger
species and more common for smaller species, particularly in environments
where resources are limiting.%, and
%2) the decreasing distance to the Hopf bifurcation with lower body sizes is suggestive of a dynamic barrier to the mass of endothermic organisms.

Previous studies have used allometric constraints to explain the periodicity
of cyclic populations
\cite{CalderIII:1983jd,Peterson:1984hj,Krukonis:1991fk}, suggesting a period
$\propto M^{0.25}$, however this relation seems to hold only for some species
\cite{Hendriks:2012fc}, and potential drivers range from predator and/or prey lifespan to competitive dynamics~\cite{Kendall:1999iy,Hogstedt:2005cr}.
%It is more difficult to ascertain whether a population is truly cycling according to an internal or external dynamic~\cite{Kendall:1999iy} or if fluctuations are due to stochasticity, and this is particularly true for longer-lived organisms.
Statistically significant support for the existence of population cycles
among mammals is predominantly based on time series for small
mammals~\cite{Kendall:1998hl}, where our model would predict more pronounced transient dynamics, given how close these points are to
the Hopf bifurcation.  On the other hand, the longer gestational times and
the increased difficulty in measurements, precludes obtaining similar-quality
data for larger organisms.
\vspace{6mm}
%, we acknowledge that absence of evidence for cyclic dynamics among longer-lived species is not necessarily evidence of absence.
 %, emphasizing the role of starvation on dynamics.

%Extinction to the left and the right
\noindent {\bf Extinction risk}\\
Within our model, higher rates of starvation result in a larger flux of the
population to the hungry state.  In this state reproduction is absent, thus
increasing the likelihood of extinction.  However, from the perspective of
population survival, it is the rate of starvation relative to the rate of
recovery that determines the long-term dynamics of the system
(Fig.~\ref{fig:fp}).  We now examine the competing effects of cyclic dynamics
vs.\ changes in steady state density on extinction risk as a function of the
ratio $\sigma/\rho$.  To this end, we computed the probability of extinction,
where extinction is defined as the population trajectory falling below
$0.2~\times$ the allometrically constrained steady state for all times
between $10^6$ and $\leq 10^8$.  This procedure is repeated for 1000
replicates of the continuous-time system shown in Eq.~\ref{eq:system} for an
organism of $M=100$ grams.  
%the initial condition isdistributed around the steady state (Eq.~\ref{eq:ss}).  Specifically
In each replicate the
initial densities are chosen to be $A(F^*,H^*,R^*)$, with $A$ a random
variable that is uniformly distributed in $[0,2]$.  By allowing the rate of
starvation to vary, we assessed extinction risk across a range of values of
the ratio $\sigma/\rho$ varying between ca. $10^{-3}$ to $5$, thus examining a
horizontal cross-section of Fig.~\ref{fig:fp}.  
As expected, higher rates of
extinction correlated with both low and high values of $\sigma/\rho$.  For
low values of $\sigma/\rho$, the increased extinction risk results from
transient cycles with larger amplitudes as the system nears the Hopf
bifurcation (Fig.~\ref{fig:ext}).  For large values of $\sigma/\rho$, higher
extinction risk arises because of to the decrease in the steady state
consumer population density.  This interplay creates an `extinction refuge'
as shown in Fig.~\ref{fig:ext}, such that for a relatively constrained range
of $\sigma/\rho$, extinction probabilities are minimized.

We find that the allometrically constrained values of $\sigma/\rho$ (with
$20\%$ variability around energetic parameter means) fall within the
extinction refuge. These values are close enough to the Hopf bifurcation to
avoid low steady state densities, and far enough away to avoid
large-amplitude transient cycles. The fact that allometric values of $\sigma$
and $\rho$ fall within this relatively small window supports the possibility
that a selective mechanism has constrained the physiological conditions that
drive starvation and recovery rates within populations.  Such a
mechanism would select for organism physiology that generates appropriate
$\sigma$ and $\rho$ values that avoid extinction.  This selection could occur
via the tuning of body fat percentages, metabolic rates, and biomass
maintenance efficiencies.
% OLD WORDING Such a mechanism would involve a feedback between the dynamics of the population and the fitness of individuals within the population, though to what extent the dynamics of the population influence rates of starvation and recovery would also involve potential tradeoffs in reproduction and somatic maintenance.
To summarize, our finding that the allometrically-determined parameters fall within this low extinction probability region suggests that the NSM dynamics may both drive---and constrain---natural animal populations.\\

@@

%Minimal mass bound
%A theoretical lower bound on mammalian body size occurs when $\epsilon=0$ where mammals have no metabolic reserves and immediately starve; this occurs at a size of $M=0$, which is obviously infeasible. 
%Such prediction gives an extreme limit on size but does not account for the subtleties of starvation dynamics
%that may limit body size (among other factors). 
The NSM correctly predicts that species with
smaller masses have larger steady-state population densities (Fig.~\ref{fig:mass}a).  However we
observe that there is a sharp change in the mass dependence of both the
steady-state densities and $\sigma/\rho$ at $M \approx 0.3$ grams
(Fig.~\ref{fig:mass}b,c).  The dependence of the rates of starvation and
recovery explain this phenomenon.  As the mass decreases, the rate of
starvation increases, while the rate of recovery declines super-exponentially.  This decline in $\rho$ occurs when the
percentage of body fat is
$1 - \big(a/b\big)^{-1/\left(\eta-1\right)}M^{-1} \approx
2\%$,
whereupon consumers are unable to recover from the hungry state.  Compellingly, this dynamic bound determined by the rate of
energetic recovery is close to the minimum observed mammalian body size of
ca. 1.3-2.5 grams (Fig.~\ref{fig:mass}b,c), a range that occurs as the
recovery rate begins its decline.  In addition to known transport
limitations~\cite{West:2002ud}, we suggest that an additional constraint of
lower body size stems from the dynamics of starvation.  This result mirrors other efforts where coincident limitations seem to constrain the
smallest possibilities for life within a particular class of
organisms~\cite{Kempes:2012hy}.

%Maximum mass bound
%Although there are upper bounds to the rate equations (e.g. when percent body fat becomes unity), they are not biologically feasible and we do not discuss them further.
A theoretical upper bound on mammalian body size is given by $\epsilon=0$, where mammals are entirely composed of metabolic reserves, and this occurs at a size of $M=8.3\times 10^8$, or $4.5\times$ the mass of a blue whale.
We determine a more realistic upper bound to body mass by assessing the
susceptibility of an otherwise homogeneous population to invasion by a
mutated subset of the population (denoted by ${}^\prime$) where individuals
have a modified proportion of body fat $M^\prime=M(1+\chi)$ where
$\chi \in [-0.5,0.5]$, thus altering the rates of starvation $\sigma$,
recovery $\rho$, and maintenance $\beta$.  There is no internal fixed point
that correspond to a state where both original residents and invaders coexist
(except for the trivial state $\chi=0$).  To assess the susceptibility to
invasion as a function of the invader mass, we determine which consumer has a
higher steady-state density for a given value of $\chi$.  We find that for
$1\leq M<10^6$g, having additional body fat ($\chi > 0$) results in a higher
steady-state invader population density
($H^{\prime *}+F^{\prime *}>H^*+F^*$).  Thus the invader has an intrinsic
advantage over the resident population.  However, for $M>10^6$, leaner
individuals ($\chi < 0$) have the advantage, and this is due to the changing
covariance between energetic rates as a function of modified energetic
reserves \sid{I don't understand the phrase after the comma}.
